---
title: "Machine Learning Course Project"
output:
  html_document:
    keep_md: yes
date: "January 27, 2016"
---

##Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: <http://groupware.les.inf.puc-rio.br/har> (see the section on the Weight Lifting Exercise Dataset).

##Data

The training data for this project are available here:

<https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv>

The test data are available here:

<https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv>
For ease of processing files and simplification of code, download the csv files and save to the set working directory, MachineLearning on the desktop.

```{r setWorkingDirectory}
setwd("~/Desktop/coursera/MachineLearning")
```

```{r loadLibraries}
## Load the preferred libraries 
library(caret) 
library(rpart)
library(randomForest)
library(AppliedPredictiveModeling)
```

Two sets of data were provided, training and test. 

The Test set contains the exercise readings for 20 participants without the "classe" variable of exercise provided. We will attempt to determine this "classe" with the use of predictive modelling, built using the training set. We will set the Test set to the side until the models are completed.
 
 With the following code, we'll read the data.

```{r readData}
pmlTrain <- read.csv("pml-training.csv", header=TRUE, na.strings=c("NA","#Div/0!"))  ## The training set
pmlTest <- read.csv("pml-testing.csv", header=TRUE, na.string=c("NA", "#Div/0!")) ## The test set -set aside
dim(pmlTrain)
dim(pmlTest)
summary(pmlTrain$classe)
```

There are 19622 records with 159 variables (the first column is just a numeric count of observations). The "classe" variable we are solving for is divided among 5 classes. 

###Cleaning the data

We'll clean out the <a href=http://www.inside-r.org/packages/cran/caret/docs/nearZeroVar>NearZeroVariance</a> variables and remove them from our data, and remove the first column, a count of the observations.
```{r cleanData}
nzv <- nearZeroVar(pmlTrain, saveMetrics=TRUE)  ## remove nearZeroVariances
pmlTrain <- pmlTrain[,nzv$nzv==FALSE]
pmlTrain <- pmlTrain[c(-1)]  ## remove first column (count)
```

Remove observations with 75% NA:
```{r}
noNAs<- pmlTrain ## find and remove 75% of NAs
for(i in 1:length(pmlTrain)) {
    if( sum( is.na( pmlTrain[, i] ) ) /nrow(pmlTrain) >= 0.75) {
        for(j in 1:length(noNAs)) {
            if( length( grep(names(pmlTrain[i]), names(noNAs)[j]) ) == 1)  {
                noNAs <- noNAs[ , -j]
            }   
        } 
    }
}
pmlTrain <- noNAs ## set back to name
rm(noNAs) ## remove excess data
dim(pmlTrain)
```
This brings us down to 58 columns.

### Split data

Now we'll split the data into a 60/40 training/test set to train the model then test the model before using for our prediction on the 20 observations in the final set. 

```{r splitData}
inTrain <- createDataPartition(y=pmlTrain$classe,p=.60,list=FALSE)
train <-pmlTrain[inTrain,]
test <- pmlTrain[-inTrain,]
```
```{r check}
dim(train)
dim(test)
```


###Prediction with Random Forests
For <a href=http://www.statmethods.net/advstats/cart.html>Random Forest information</a>
```{r randomForest}
set.seed(4726)
modelFitA1 <- randomForest(classe ~ ., data=train)
predictionA1 <- predict(modelFitA1, test, type = "class")
modelRF <- confusionMatrix(predictionA1, test$classe)
modelRF
plot(modelFitA1)
importance(modelFitA1)
print(modelRF)
```


###Prediction with Decision Trees
<a href=http://blog.revolutionanalytics.com/2013/06/plotting-classification-and-regression-trees-with-plotrpart.html>Create decision tree</a>

Instead of plotting a decision tree, we can quickly look at a graph of the cross-validation results, and see that the error rate is higher than the random forest method. 

```{r decisionTree}
set.seed(4726)
modelFitB1 <- rpart(classe ~., method="class", data=train)
plot(modelFitB1)
text(modelFitB1)
plotcp(modelFitB1)
```

A quick look at the results on teh decision tree and we see the variables we're looking for intermingled. The Random Forest has a 99.83% accuracy rate, a .17% out-of-sample error rate, so we'll progress with this for our prediction set.

###Predicting our results
First, we'll use the same cleaning methods as above:
```{r cleanTestData}
cleanFormat <- colnames(train[,-58]) # classe column removal
pmlTest <-pmlTest[cleanFormat]
dim(pmlTest)
```

And we'll coerce the data into the same format:

```{r coerceData}
for (i in 1:length(pmlTest) ) {
    for(j in 1:length(train)) {
        if( length( grep(names(train[i]), names(pmlTest)[j]) ) == 1)  {
            class(pmlTest[j]) <- class(train[i])
        }      
    }      
}

# To get the same class between pmlTest and train
pmlTest <- rbind(train[2,-58], pmlTest) ## remove excess rows
pmlTest <- pmlTest[-1,]

```

Then we apply the prediction model to the data:
```{r predictingFinal}
predictionFinal <-predict(modelFitA1, pmlTest, type="class")
```

And our final results.
```{r finalResults}
predictionFinal
```


